---
title: 'More control on the output of **tradeSeq**'
author: "Koen Van den Berge and Hector Roux de Bézieux"
date: "25/02/2020"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Finer settings for tradeSeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE}
library(knitr)
```

# Introduction

`tradeSeq` is an `R` package that allows analysis of gene expression along trajectories. While it has been developed and applied to single-cell RNA-sequencing (scRNA-seq) data, its applicability extends beyond that, and also allows the analysis of, e.g., single-cell ATAC-seq data along trajectories or bulk RNA-seq time series datasets. For every gene in the dataset, `tradeSeq` fits a generalized additive model (GAM) by building on the `mgcv` R package. It then allows statistical inference on the GAM by assessing contrasts of the parameters of the fitted GAM model, aiding in interpreting complex datasets. All details about the `tradeSeq` model and statistical tests are described in our preprint [@VandenBerge2019a].

In this vignette, we analyze a subset of the data from [@Paul2015]. 
A `SingleCellExperiment` object of the data has been provided with the [`tradeSeq`](https://github.com/statOmics/tradeSeqpaper) package and can be retrieved as shown below. The data and UMAP reduced dimensions were derived from following the [Monocle 3 vignette](http://cole-trapnell-lab.github.io/monocle-release/monocle3/#tutorial-1-learning-trajectories-with-monocle-3). 

# Load data

```{r, warning=F, message=F}
library(tradeSeq)
library(RColorBrewer)
library(SingleCellExperiment)
library(slingshot)

# For reproducibility
RNGversion("3.5.0")
palette(brewer.pal(8, "Dark2"))
data(countMatrix, package = "tradeSeq")
counts <- as.matrix(countMatrix)
rm(countMatrix)
data(crv, package = "tradeSeq")
data(celltype, package = "tradeSeq")
```

# Choosing K: a deeper dive into the output from __evaluateK__

If you want to use the different diagnostic plots to choose the number of knots when fitting the smoothers, you can disable to plotting option

```{r, eval=FALSE}
set.seed(5)
icMat <- evaluateK(counts = counts, sds = crv, k=3:6, nGenes = 200,
                   verbose = FALSE, plot = FALSE)
print(icMat[1:2, ])
```

Then, you can explore

# Fit additive models
## Parallel computing

```{r}

```

## Convergence issues on small or zero-inflated datasets

If you're working with a dataset that has a limited number of cells, or if you
are incorporating zero inflation weights, the GAMs may be harder to fit, as
noted by the warnings when running `fitGAM`. In that case, the situation 
might improve if you allow for more iterations in the GAM fitting. This can be
done with the `control` argument of `fitGAM`.

```{r}
library(mgcv)
control <- gam.control()
control$maxit <- 1000 #set maximum number of iterations to 1K
# pass to control argument of fitGAM as below:
# 
# gamList <- fitGAM(counts = counts,
#                   pseudotime = slingPseudotime(crv, na = FALSE),
#                   cellWeights = slingCurveWeights(crv),
#                   control = control)
```

## tradeSeq list output

The output from `fitGAM` will be different if one sets `sce=FALSE`, and less memory efficient. Instead of a `SingleCellExperiment` object, it will return a list with all fitted `mgcv` models. Most functions we have discussed above work exactly the same with the list output. However, the list output functionality is a little bit bigger, and here we discuss some capabilities that are only available with the list output.

```{r}
gamList <- fitGAM(counts,
                  pseudotime = slingPseudotime(crv, na = FALSE),
                  cellWeights = slingCurveWeights(crv),
                  nknots = 6, sce = FALSE)
```


First, one may explore the results of a model by requesting its summary.

```{r}
summary(gamList[["Irf8"]])
```

Related to the `associationTest`, one can extract the p-values generated by the `mgcv` package using the `getSmootherPvalues` function. These p-values are derived from a test that assesses the null hypothesis that all smoother coefficients are equal to zero. Note, however, that their interpretation is thus more complex. A significant lineage for a particular gene might thus be the result of (a) a different mean expression in that lineage as compared to the overall expression of that gene, or (b) significantly varying expression along that lineage, even if the means are equal, or (c) a combination of both.
This function extracts the p-values calculated by `mgcv` from the GAM, and will return `NA` for genes that we were unable to fit properly.
Similarly, the test statistics may be extracted with `getSmootherTestStats`.
Since this dataset was pre-filtered to only contain relevant genes, all p-values (test statistics) will be very low (high).
Note, that these functions are only applicable with the list output of `tradeSeq`, and not with the `SingleCellExperiment` output. We will therefore not evaluate these here.

```{r, eval=FALSE}
pvalLineage <- getSmootherPvalues(gamList)
statLineage <- getSmootherTestStats(gamList)
```

# Clustering

```{r}
yhat <- predictCells(models = sce, gene = "Irf8")

ysmooth <- predictSmooth(models = sce, gene = "Irf8", nPoints = 40)
```

# Session

```{r}
sessionInfo()
```
